---
title   : 'Data Science for Linguists'
subtitle : 'The linear model: [Assumptions, diagnotics, and interpretation]{style="color: #888; font-size: 0.9em;"}'
author   : "Joseph V. Casillas, PhD"
institute: "Rutgers University<mybr>Spring 2025<br>Last update: `r Sys.Date()`"
---

```{r}
#| label: load-helpers
#| echo: false 
#| message: false 
#| warning: false

source(here::here("assets", "scripts", "helpers.R"))
read_chunk(here("05_lm", "02_assumptions_diagnostics", "index_files", "scripts", "assumptions_diagnostics.R"))
```

# {.transition visibility="uncounted"}

{{< tweet user=ChelseaParlett id=1352717277976694784 >}}
<!-- regression the movie tweet --> 


# Assumptions {.transition}

---

## Assumptions {background-image="./index_files/img/assumptions.png" background-size="350px" background-position="95% 50%" data-menu-title="General assumptions"}

### We make a lot of assumptions

::: {.columns}
::: {.column style="font-size: 0.7em; width: 75%;"}
<ru-blockquote>
[1.]{color="lightblue"} the act of assuming or taking for granted  
[2.]{color="lightblue"} a hypothesis that is taken for granted  
[3.]{color="lightblue"} a thing that is accepted as true or as certain to happen, without proof
</ru-blockquote>
:::
:::

::: {.columns}
::: {.column .closelist .absolute bottom="10%" style="font-size: 0.7em;"}
- About the world
- About others
- About reactions
- About academia
:::

::: {.column .closelist .absolute bottom="10%" right="0" style="font-size: 0.7em;"}
- The sun will rise tomorrow
- People are generally X
- They will find my joke funny
- Reviewer 2 is evil
:::
:::

---

## Assumptions {background-image="https://raw.githubusercontent.com/jvcasillas/media/master/rstats/memes/lm_assumptions.png" background-size="350px" background-position="95% 50%" data-menu-title="Statistical assumptions"}

### Statistical assumptions

::: {.columns}
::: {.column style="font-size: 0.9em; width: 75%;"}
- Every model has assumptions 
- They need to be met for the model to work properly and be trustworthy
- It is not standard practice to report whether all assumptions have been met when writing up results... [but it should be]{.emph}
- We will make a habit of assessing models and incorporating the relevant information in prose
:::
:::

---

## Assumptions {background-image="https://raw.githubusercontent.com/jvcasillas/media/master/rstats/memes/lm_assumptions.png" background-size="350px" background-position="95% 50%" data-menu-title="Regression assumptions"}

### Regression assumptions

We can break the assumptions up into 3 areas: 

1. Model specification
2. Measurement error
3. The error term



# Model specification {.transition}

---

## Model specification {data-menu-title="Specification error"}

### Specification error

There should be no specification error

- The relationship between x<sub>i</sub> and y<sub>i</sub> is linear: 
$$y_i = \alpha + \beta x_i + \epsilon _i$$
- Including irrelevant variables 🤷🏽
- Omitting relevant variables 👎🏽

::: {.fragment}
### Importance: [**HIGH**]{.emph}
:::

---

## Model specification {data-menu-title="Including an irrelevant variable" background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%"}

### Including an irrelevant variable (sin of commission): 🤷🏽

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- Similar to a Type I error
- Adds to the error of prediction, but it does not bias the parameter estimates 
- If model includes multiple predictors, other estimates are not influenced by the irrelevant variable
- Though the parameter estimates won't be biased, the standard error terms 
around each beta weight will increase (this affects t-ratio and p-value)
:::
:::

---

## {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%"}

[Model specification]{.emph .p-font style="font-size: 1.85em;"}

[Including an irrelevant variable (sin of commission): 🤷🏽]{.p-font style="font-size: 1.3em; color: #666666"}

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- For example, imagine a situation in which you model employee performance as a function of a bunch of good predictors
- Adding the variable of astrological horoscope sign will not bias the other estimates
- You are considering something that doesn't matter, i.e., astrology
:::
:::

---

## Model specification {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%" data-menu-title="Excluding a relevant variable"}

### Excluding a relevant variable (sin of omission): 👎🏽

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- Similar to a Type II error
- This is much more serious!
- This will bias your parameter estimates
- You are leaving something important out (variance left unexplained)
:::
:::

---

## {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%"}

[Model specification]{.emph .p-font style="font-size: 1.85em;"}

[Excluding a relevant variable (sin of omission): 👎🏽]{.p-font style="font-size: 1.3em; color: #666666"}

::: {.columns}
::: {.column style="font-size: 0.85em; width: 70%;"}
- For example, if you want to model RT as a function of working memory but you forget to look at age
- Then your prediction might be wrong
- You don’t know which way you are biasing things either!
- You could be either overestimating or underestimating the other regression parameters
:::
:::

---

## {background-image="./index_files/img/sin.png" background-size="400px" background-position="95% 50%" data-menu-title="What to do?"}

[Model specification]{.emph .p-font style="font-size: 1.85em;"}

[What to do]{.p-font style="font-size: 1.3em; color: #666666"}

#### [After including an irrelevant predictor:]{color="black"}

::: {.closelist style="font-size: 0.8em;"}
- Drop the irrelevant variable
- Not much of a problem
:::

<mybr>

::: {.fragment}
#### [After excluding a relevant predictor:]{color="black"}

::: {.closelist style="font-size: 0.8em;"}
- Look at data again to see if it is in there
- Hope and Pray!
- If it’s not in you data, you are SOL
- You need to repeat the study!
- There is a problem in your design and there is no mathematical solution to fix this problem
:::
:::



# Measurement error {background-image="./index_files/img/measure0.png" background-size="contain" background-position="100% 50%"}

---

## Measurement error {data-menu-title="Measuring is hard" background-image="https://media.tenor.com/AFJK9BNwiZIAAAAM/harry-styles-tape-measure.gif" background-size="300px" background-position="98% 10%"}

### Measuring things is hard

::: {.columns}
::: {.column .fragment .absolute left="0" top="35%" width="55%"}
- How far away is that lion?
- How much milk do I need?
- Can I park there?
- How much weight did I gain?
:::

::: {.column .absolute right="0" top="30%" width="10%"}
:::

::: {.column .fragment .absolute right="0" top="35%" width="35%"}
- Human error
- Specificity
- Operationalizations
- Repeatability
:::
:::

---

## Measurement error {data-menu-title="Minimize measurement error" background-image="./index_files/img/measure1.png" background-size="contain" background-position="110% 50%"}

### There should be no measurement error

::: {.columns}
::: {.column .absolute top="35%"}
- Variables x<sub>i</sub> and y<sub>i</sub> should be as accurate as possible
:::
:::

::: {.absolute bottom="5%"}
[Importance: [high]{.emph}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}
:::

---

## Measurement error {data-menu-title="Consequences" background-image="https://media4.giphy.com/media/v1.Y2lkPTc5MGI3NjExNmV2M2xiZXBhemV3MGdhNzFreWpuNXg1Mmd5YnB4MDh0bnRwNWhnZyZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/k5lj4s1qxaSyI/giphy.webp" background-size="500px" background-position="98% 50%"}

### Consequences of measurement error

::: {.columns}
::: {.column .incremental style="width: 65%; font-size: 0.9em;"}
- Your models will estimate the parameters you specify
- But you won't know if the estimates are reliable
- Large degree of unquantifiable uncertainty
- Cannot "safely" do NHST
- Your research may be reproducible, but probably will not be replicable
:::
:::

<!-- anchorman gif, 60% of the time it works every time -->

---

## Measurement error {data-menu-title="What to do?" background-image="./index_files/img/measure2.gif" background-size="500px" background-position="98% 50%"}

### What you can do

::: {.columns}
::: {.column .incremental style="width: 65%; font-size: 0.9em;"}
- The solution to this problem revolves around proper planning a priori
- Operationally define every feasible aspect of *how* and *what* you will measure
  - Use previous literature, if available
  - Preregister decisions
- Repeated measures, if possible
- Random, 3rd party quality checks
- Measurement error models, if possible
:::
:::

<!-- anchorman gif, 60% of the time it works every time -->



# The error term {.transition}

---

## The error term {data-menu-title="About the error term"}

### About the error term

The error term should meet the following characteristics...

- Mean = 0
- Homoskedasticity
- No autocorrelation
- Predictor not correlated with error term
- Error is normally distributed 

---

## About the error term {data-menu-title="Mean of 0"}

### Mean of 0

If the mean of the error term deviates far from 0...

- Intercept can be biased
- Slope estimates will not be biased

[Importance: [low]{color="blue"}... unless you are interested in the intercept.]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## About the error term {data-menu-title="Homoskedasticity"}

### Homoskedasticity

::: {.closelist}
- Variance around predicted values should be consistent
- Common simple inspection is to look at scatter plot of fitted vs. x-values (should look like blob with no interesting patterns)
- If variance is heteroskedastic you will typically see fan-like patterns
- Will not bias parameter estimates
- Will increase confidence intervals, thus affects t-ratios and p-values
:::

[Importance: [medium/high]{color="orange"}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## {background-color="black" visibility="uncounted" .center}

<center>
<iframe height="390" width= "520" src="https://www.tiktok.com/player/v1/7076276399505558827?music_info=1&description=1" allow="fullscreen" title="test"></iframe>
</center>

::: footer
<https://www.tiktok.com/@chelseaparlettpelleriti/video/7076276399505558827>
:::

<!-- homoskedasticity -->

---

## About the error term {data-menu-title="Autocorrelation"}

### No autocorrelation

::: {style="font-size: 0.8em;"}
- If the residuals are autocorrelated, it means that the prediction error of a given observation depends on that of the previous observation
- This shows up as a clear unexplained pattern in the y variable 
- Most common in repeated measures and longitudinal data
- Will not bias parameter estimates
- Will affect confidence intervals, t-ratios, p-values
- Increased chance of Type II error
:::

[Importance: [High]{.emph}, but uncommon in standard regression]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## About the error term {data-menu-title="Predictor correlation with error term"}

### Predictor(s) should not be correlated with error term

- Typically the result of omitting a relevant variable (sin of omission)
- Will bias parameter estimates
- Solution: include missing variable

[Importance: [high]{.emph}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}

---

## About the error term {data-menu-title="Normally distributed errors"}

### Error (residuals) should be normally distributed

::: {.incremental .closelist style="font-size: 0.8em;"}
- There is no a priori reason for error to be anything but normally distributed 
- It should become standard practice for you to examine your residuals to see if they are normally distributed or not
- If they aren't normally distributed there is a substantial possibility of Type II error
- If this is the case, then the residuals may not be "pure" error, and you may have omitted a relevant variable from the equation that is making the residuals not be normally distributed
- In other words, the residuals may contain systematic variance that can still be explained by something else
- You cannot conclude with 100% certainty that a Type II error has been 
committed, but you might strongly suspect it
:::

[Importance: [high]{.emph}]{.p-font .fragment style="font-size: 1.2em; color: #666666;"}







# Summing up {.smaller}

::: {.closelist}
- Model specification
  - The relationship between x<sub>i</sub> and y<sub>i</sub> is linear
  - Including irrelevant variables 
  - Omitting relevant variables
:::

- Measurement error

::: {.closelist}
- The error term
  - Mean = 0
  - Homoskedasticity
  - No autocorrelation
  - Predictor not correlated with error term
  - Error is normally distributed
:::




# Diagnostics Examples {.transition}

---

## Diagnostics {.smaller data-menu-title="More about residuals"}

### Remember those "residuals"?

::: {.closelist}
- A residual represents prediction error in our model with regard to a single point. 
  - Our `mtcars` model predicted that a 6 ton car should get 5.22 mpg. 
  - If in reality it gets 10.22 mpg, then the residual for that specific data point would be 5 mpg. 
:::

::: {.fragment}
- Remember: All models are wrong. 
:::

::: {.closelist .fragment}
- Our models will always be off by at least a little bit for most of the observations we try to fit. 
- Nonetheless, it is good practice to examine the residuals of your models in order to help assess goodness-of-fit.
- Specifically, we need to make sure that there aren't any unexpected patterns because that would suggest that our model does not properly fit our data.
:::

---

## {background-image="./index_files/img/sick-computer.png" background-size="400px" background-position="95% 50%"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[Considerations for model diagnostics]{.p-font style="font-size: 1.2em; color: #666666"}

<br>

[1. Model assumptions]{.p-font style="font-size: 1.2em;"}

[2. Outliers]{.p-font style="font-size: 1.2em;"}

---

## {data-menu-title="Assumption of linear relationship" .smaller}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### The relationship between x<sub>i</sub> and y<sub>i</sub> is linear.

1. Double check linear specification
2. Eyeball it

```{r}
#| label: fit_mods
data(assumptions_data)
mod1 <- lm(y ~ x, data = assumptions_data)
mod2 <- lm(y_quad ~ x, data = assumptions_data)
```

<mybr>

::: {.columns}
::: {.column}
```{r}
#| label: assumptions_linear_plot
#| fig-asp: 0.8
assumptions_data |> 
  ggplot() + 
  aes(x = x, y = y) + 
  geom_point(
    pch = 21, color = 'darkgrey', fill = 'darkred', 
    stroke = 1, size = 4
  ) + 
  ds4ling_bw_theme(base_size = 24)
```
:::

::: {.column}
```{r}
#| label: assumptions_quadratic_plot
#| fig-asp: 0.8
assumptions_data |> 
  ggplot() + 
  aes(x = x, y = y_quad) + 
  geom_point(
    pch = 21, color = 'darkgrey', fill = 'darkred', 
    stroke = 1, size = 4
  ) + 
  ds4ling_bw_theme(base_size = 24)
```
:::
:::

---

## {data-menu-title="Mean of residuals = 0" .smaller}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### The mean of residuals is zero

1. Fit model
2. Get residuals
3. Test manually

<mybr>

::: {.columns}
::: {.column}
```{r}
#| label: model-ex-code1
#| echo: true
# Linear relationship
# Fit model and print summary
mod1 <- lm(y ~ x, data = assumptions_data)
summary(mod1$residuals)
```
:::

::: {.column}
```{r}
#| label: model-ex-code2
#| echo: true
# Non-linear relationship
# Fit model and print summary# Fit mod
mod2 <- lm(y_quad ~ x, data = assumptions_data)
summary(mod2$residuals)
```
:::
:::

---

## {data-menu-title="Homoskedasticity of residuals"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### Homoskedasticity of residuals

```{r}
#| label: homoskedasticity_linear
#| fig-asp: 0.45
autoplot(mod1, which = c(1, 3)) + 
  ds4ling_bw_theme(base_size = 18)
```

::: aside
mod1
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[Homoskedasticity of residuals]{.p-font style="font-size: 1.2em; color: #666666"}

```{r}
#| label: homoskedasticity_quadratic
#| fig-asp: 0.45
autoplot(mod2, which = c(1, 3)) + 
  ds4ling_bw_theme(base_size = 18)
```

::: aside
mod2
:::

---

## {data-menu-title="No autocorrelation of residuals"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### No autocorrelation of residuals

::: {.columns}
::: {.column}
1. Visual inspection
2. Durbin-Watson test

```{r}
#| label: linear_durbin_watson
#| comment: ''
# formal test: Durbin-Watson test
lmtest::dwtest(mod1)
```
:::

::: {.column}
```{r}
#| label: linear_autocorrelation
#| fig-asp: 0.8
#| out-extra: "style=float:right"
# visual inspection
acf(mod1$residuals) 
```
:::
:::

::: aside
mod1
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[No autocorrelation of residuals]{.p-font style="font-size: 1.2em; color: #666666;"}

::: {.columns}
::: {.column}
1. Visual inspection
2. Durbin-Watson test

```{r}
#| label: quadratic_durbin_watson
#| comment: ''
# formal test: Durbin-Watson test
lmtest::dwtest(mod2)
```
:::

::: {.column}
```{r}
#| label: quadratic_autocorrelation
#| fig-asp: 0.8
# visual inspection
acf(mod2$residuals) 
```
:::
:::

::: aside
mod2
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[No autocorrelation of residuals - Correction]{.p-font style="font-size: 1.2em; color: #666666;"}

::: {.columns}
::: {.column .small-code}
```{r}
#| label: auto_cor_fix
#| fig-asp: 0.8
#| echo: true
autocor_data <- data.frame(
  assumptions_data, 
  resid_mod2 = mod2$residuals
)

autocor_data1 <- DataCombine::slide(
  autocor_data, 
  Var = "resid_mod2", 
  NewVar = "lag1", 
  slideBy = -1
)
autocor_data2 <- na.omit(autocor_data1)
mod2_fix <- lm(y_quad ~ x + lag1, data = autocor_data2)
```

```{r}
#| label: autocor-durbin-watson
#| comment: ''
lmtest::dwtest(mod2_fix)  # formal test: Durbin-Watson test
```
:::

::: {.column}
```{r}
#| label: auto-cor-fix-plot
#| fig-asp: 0.8
acf(mod2_fix$residuals)
```
:::
:::

::: footer
mod2
:::

---

## {data-menu-title="Predictors and residuals are uncorrelated"}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### Predictors and residuals are uncorrelated

::: {.closelist .small-code}
- Test for correlation
- Think about your study

```{r}
#| label: pred-resid-corr-test-linear
# do correlation test 
cor.test(assumptions_data$x, mod1$residuals)
```
:::

::: aside
mod1
:::

---

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

[Predictors and residuals are uncorrelated]{.p-font style="font-size: 1.2em; color: #666666;"}

::: {.closelist .small-code}
- Test for correlation
- Think about your study

```{r}
#| label: pred_resid_corr_test_quadratic
# do correlation test 
cor.test(assumptions_data$x, mod2$residuals)
```
:::

::: aside
mod2
:::

---

## {data-menu-title="Normality of residuals" .smaller}

[Diagnostics]{.emph .p-font style="font-size: 1.75em;"}

### Normality of residuals

- QQplots

```{r}
#| label: normal_resids_linear-quadratic
#| fig-asp: 0.45
#| fig-align: 'center'
ap1 <- autoplot(mod1, which = 2) + 
  ds4ling_bw_theme(base_size = 18)

ap2 <- autoplot(mod2, which = 2) + 
  ds4ling_bw_theme(base_size = 18)

ap1 + ap2
```





# Dealing with influential data points {data-menu-title="Influential data points" .center .transition}

---

## Outliers/influential data points {.smaller}

- An influential point is one that would significantly change the fit if 
removed from the data 
- [Cook's distance]{color="blue"} is a commonly used influence measure 

. . .

**Leverage**

- The leverage of an observation measures its ability to move the regression 
line by simply moving up/down along the y-axis.
- The measurement represents the amount by which the predicted value would 
change if the observation was shifted one unit in the y-direction.
- The leverage always takes values between 0 and 1.
- A point with 0 leverage does not effect the regression line.

---

[Influential data points]{.emph .p-font style="font-size: 1.75em;"}

<br>

::: {.columns}
::: {.column}

```{r}
#| label: cooks_distance_linear
#| fig-asp: 0.9
# autoplot(mod2, which = 4)
# Cook's D plot
cutoff <- function(model) {
  # identify D values > 4 / (n-k-1)
  n <- length(model$residuals)
  k <- length(model$coefficients)
  cutoff <- 4 / (n - k - 1)
  return(cutoff)
}

tibble(
  observation = as.numeric(names(cooks.distance(mod1))), 
  cooksd = cooks.distance(mod1)
  ) |>
  arrange(observation) |> 
  mutate(
    label = if_else(
      condition = cooksd > cutoff(mod1), 
      true = as.character(observation), 
      false = ' '
    )
  ) |> 
ggplot() + 
aes(x = observation, y = cooksd, label = label) + 
geom_hline(yintercept = cutoff(mod1), lty = 3) + 
geom_point() + 
geom_text(nudge_y = 0.1) + 
geom_segment(aes(xend = observation, y = 0, yend = cooksd)) + 
labs(
  x = "Obs. Number", 
  y = "Cook's distance", 
  title = "Cook's distance"
) + 
ds4ling_bw_theme(base_size = 22)
```

:::

::: {.column}

```{r}
#| label: cooks_distance_quadratic
#| fig-asp: 0.9
tibble(
  observation = as.numeric(names(cooks.distance(mod2))), 
  cooksd = cooks.distance(mod2)
  ) |>
  arrange(observation) |> 
  mutate(
    label = if_else(
      condition = cooksd > cutoff(mod2), 
      true = as.character(observation), 
      false = ' '
    )
  ) |> 
  ggplot() +
  aes(x = observation, y = cooksd, label = label) + 
  geom_hline(yintercept = cutoff(mod2), lty = 3) + 
  geom_point() + 
  geom_text(nudge_y = 0.05) + 
  geom_segment(aes(xend = observation, y = 0, yend = cooksd)) + 
  labs(
    x = "Obs. Number", 
    y = "Cook's distance", 
    title = "Cook's distance"
  ) + 
  ds4ling_bw_theme(base_size = 22)
```
:::
:::


---

[Influential data points]{.emph .p-font style="font-size: 1.75em;"}

```{r}
#| label: leverage_linear_quadratic2
#| fig-asp: 0.5
ap3 <- autoplot(mod1, which = 5) + 
  ds4ling_bw_theme(base_size = 18)

ap4 <- autoplot(mod2, which = 5) + 
  ds4ling_bw_theme(base_size = 18)

ap3 + ap4
```


---

## 

[Influential data points]{.emph .p-font style="font-size: 1.75em;"}

::: {.columns}
::: {.column}

<br>

```{r}
#| label: mod1_with_outlier
#| fig-asp: 0.9
noOut <- lm(y ~ x, data = assumptions_data[1:50, ])

assumptions_data |> 
  mutate(
    observation = row_number(),
    label = if_else(
      condition = observation == 51, 
      true = "51", 
      false = " "
    )
  ) |> 
  ggplot() + 
  aes(x = x, y = y, label = label) + 
  geom_point(aes(color = label), size = 3, show.legend = FALSE) + 
  geom_text(nudge_y = 0.3, size = 5, color = "darkred") + 
  geom_smooth(method = lm, se = F, size = 1.5, color = "darkred", 
    formula = "y ~ x") + 
  geom_abline(intercept = noOut$coef[1], slope = noOut$coef[2], 
              color = 'blue', size = 1.5) + 
  scale_color_brewer(palette = "Set1", guide = FALSE, direction = -1) + 
  ds4ling_bw_theme(base_size = 22)
```
:::


::: {.column}

#### **Mod1 with influential data point**:

::: box-error

```{r}
#| label: model_print_with_outlier
#| results: 'asis'
b_with <- coef(mod1)
cat(sprintf("$$y = %.02f + %.02f x$$", b_with[1], b_with[2]))
```
:::

<br>

#### [Mod1 without influential data point]{color="blue"}: 

::: box-note

```{r}
#| label: model_print_without_outlier
#| results: 'asis'
b_without <- coef(noOut)
cat(sprintf("$$y = %.02f + %.02f x$$", b_without[1], b_without[2]))
```
:::
:::
:::

---

## Global test of model assumptions

- It is also possible to use the package `gvlma` to test model assumptions. 
- It seems rather conservative
- I don't know too much about it. 

---

```{r}
#| label: gvlma_test_with
#| comment: "##"
library("gvlma")
gvmodel_1with <- gvlma(mod1)
gvmodel_1with
```

---

```{r}
#| label: gvlma_test_without
#| comment: "##"
gvmodel_1_without <- gvlma(noOut)
gvmodel_1_without 
```

---

```{r}
#| label: gvlma_test_mod2
#| comment: "##"
gvmodel_2 <- gvlma(mod2)
gvmodel_2
```

---

## {.center}

<iframe src="https://gallery.shinyapps.io/slr_diag/" style="border:none;" height="600" width="1300"></iframe>

<!-- assumptions shiny app -->




# Interpretation {.transition}

---

## Interpretation {.smaller data-menu-title="Make sense of it all"}

### Make sense of it all

::: {.columns}
::: {.column width="60%"}
```{r}
#| label: model_output
#| comment: ''
cars_mod <- lm(mpg ~ wt, data = mtcars)
summary(cars_mod)
```
:::

::: {.column width="40%"}

::: {style="font-size: 0.8em;"}
1. Function call
2. Model residuals
3. Model coefficients
4. Significance codes
5. Variance explained
6. F-ratio
:::
:::
:::

---

## Interpretation {.smaller data-menu-title="Function call"}

### Make sense of it all

::: {.columns}
::: {.column width="60%"}

```{r}
#| label: model_output1
#| comment: ''
summary(cars_mod)$call
```
:::

::: {.column width="40%"}
::: {style="font-size: 0.8em;"}
1. Function **[call]{.emph}**
2. Model residuals
3. Model coefficients
4. Significance codes
5. Variance explained
6. F-ratio

::: box-note
- This is the model you fit using `lm()`. 
- It is the exact same code you typed into R. 
- Use to double check to see if you made any typos.
:::
:::
:::
:::


---

## Interpretation {.smaller data-menu-title="Model residuals"}

### Make sense of it all

::: {.columns}
::: {.column width="60%"}

```{r}
#| label: model_output2
#| comment: ''
#| results: 'hold'
summary(cars_mod$residuals)
```
:::

::: {.column width="40%"}
::: {style="font-size: 0.8em;"}
1. Function call
2. Model [**residuals**]{.emph}
3. Model coefficients
4. Significance codes
5. Variance explained
6. F-ratio

::: box-note
- Should be normally distributed
- Absolute values of 1Q/3Q should be similar
- Mean = 0, median close to 0
- If anything is off you can see it here, but check residual plots
:::
:::
:::
:::


---

## Interpretation {.smaller data-menu-title="Model coefficients"}

### Make sense of it all

::: {.columns}
::: {.column width="55%"}
```{r}
#| label: model_output3
#| comment: ''
#| results: 'hold'
summary(cars_mod)$coef
```
:::

::: {.column width="45%"}
::: {style="font-size: 0.8em;"}
1. Function call
2. Model residuals
3. Model [**coefficients**]{.emph}
4. Significance codes
5. Variance explained
6. F-ratio

::: box-note
- Meat and potatoes of the output
- Parameter estimates of intercept and predictor(s)
- Express strength of relationship between predictor(s) and criterion
- One unit change in predictor will change outcome by...
:::
:::
:::
:::


---

## Interpretation {.smaller data-menu-title="Significance™"}

### Make sense of it all

::: {.columns}
::: {.column width="60%"}
```{r}
#| label: model_output4
#| comment: ''
#| results: 'hold'
summary(cars_mod)
```
:::

::: {.column width="40%"}
::: {style="font-size: 0.8em;"}
1. Function call
2. Model residuals
3. Model coefficients
4. [**Significance**]{.emph} codes
5. Variance explained
6. F-ratio

::: box-note
- Assessment of statistical significance of predictor(s) and the intercept 
- Everything except "." is probably "good"
:::
:::
:::
:::


---

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[Statistical Significance]{.p-font style="font-size: 1.2em; color: #666666"}

- As long as the effect is not statistically equivalent to 0 it is called 
statistically significant

- It may be an effect of trivial magnitude

- Basically, it means that this prediction is better than nothing

- It doesn’t really mean it is really "significant" in the terms that we think of as significance

- It doesn’t indicate importance


---

## {.smaller}

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[Significance versus Importance]{.p-font style="font-size: 1.2em; color: #666666"}

**How do we know if the "significant" effect we found is actually important?**

- The coefficient of determination (r-squared) tells you how much of the variance you have explained
- It tells us how big the effect is and not just that it is not equal to zero
- You want to know if your predictions are better than chance alone (F-ratio) but you also want to know how explanatory your predictions are (r-squared)

. . .

**How important are the chosen predictors?**

- The numerators in both ratios are similar
- They represent the predicted portion of the variance


---

## Interpretation {.smaller data-menu-title="Variance explained"}

### Make sense of it all

::: {.columns}
::: {.column width="60%"}
```{r}
#| label: model_output5
#| comment: ''
#| results: 'hold'
summary(cars_mod)
```
:::

::: {.column width="40%"}
::: {style="font-size: 0.8em;"}
1. Function call
2. Model residuals
3. Model coefficients
4. Significance codes
5. [**Variance explained**]{.emph}
6. F-ratio

::: box-note
- Assessment of R<sup>2</sup>
- More variance explained is better
- Should pretty much always be reported
:::
:::
:::
:::

---

## Interpretation {.smaller data-menu-title="F-ratio"}

### Make sense of it all

::: {.columns}
::: {.column width="60%"}
```{r}
#| label: model_output6
#| comment: ''
#| results: 'hold'
summary(cars_mod)
```
:::

::: {.column width="40%"}
::: {style="font-size: 0.8em;"}
1. Function call
2. Model residuals
3. Model coefficients
4. Significance codes
5. Variance explained
6. [**F-ratio**]{.emph}

::: box-note
- Assesses overall significance... omnibus model
- If F-ratio is significant, at least one predictor or intercept is too
- If F-ratio is not significant your experiment is over
:::
:::
:::
:::


---

## {.smaller}

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[A note about F-ratios]{.p-font style="font-size: 1.3em; color: #666666"}

**If you have an ANOVA background... Mean Squared Deviations**

$$MS_{Total} = \frac{\sum{(y_i - \bar{y})^2}}{n-1}$$  

$$MS_{Predicted} = \frac{\sum{(\hat{y}_i - \bar{y})^2}}{k}$$  

$$MS_{Error} = \frac{\sum{(y_i - \hat{y}_i)^2}}{n - k - 1}$$

::: box-tip
$$F_{(k),(n-k-1)} = \frac{\sum{(\hat{y}_i - \bar{y})^2} / (k)}{\sum{(y_i - \hat{y}_i)^2} / (n - k - 1)}$$
:::


---

## {.smaller}

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[A note about F-ratios]{.p-font style="font-size: 1.3em; color: #666666"}

**If you have an ANOVA background... Mean Squared Deviations**

$$MS_{Total} = \frac{SS_{Total}}{df_{Total}}$$

$$MS_{Predicted} = \frac{SS_{Predicted}}{df_{Predicted}}$$

$$MS_{Error} = \frac{SS_{Error}}{df_{Error}}$$

::: box-tip
$$F = \frac{MS_{Predicted}}{MS_{Error}}$$
:::


---

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[Degrees of Freedom]{.p-font style="font-size: 1.3em; color: #666666"}

[**Derived from the number of sample statistics used in your computation:**]{style="font-size: 0.8em;"}

- e.g., for a standard deviation, you subtract all the raw scores from the mean
- Since you used the sample mean, you used up 1 df
- df = n - 1


---

## {.smaller}

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[Degrees of Freedom]{.p-font style="font-size: 1.3em; color: #666666"}

[**In the denominator of the standard deviation, you are using a sample statistic, not a population parameter, so you have df = n - 1:**]{style="font-size: 0.8em;"}

- n - 1 reflects the fact that you used a statistic and if you know one number 
(the mean) there is less uncertainty remaining
- If you know the mean and you have 10 scores, then you only need 9 of the 
remaining scores to predict all 10 of them

[**Usually we have an F-table with associated degrees of freedom to show us whether the F-ratio is "statistically significant":**]{style="font-size: 0.8em;"}

- Numerator df = k 
- Denominator df = n - k - 1


---

[Interpretation]{.emph .p-font style="font-size: 1.75em;"}

[Degrees of Freedom]{.p-font style="font-size: 1.3em; color: #666666"}

[$$df_{Total} = n - 1$$]{style="color: #555555"}

[$$df_{Predicted} = k$$]{style="color: #555555"}

[$$df_{Error} = n - k - 1$$]{style="color: #555555"}

[$$df_{Total} = df_{Predicted} + df_{Error}$$]{style="font-size: 1.1em;"}




# [Exercises](https://www.ds4ling.jvcasillas.com/slides/05_lm/02_assumptions_diagnostics/index_files/scripts/assumptions_walkthrough_ex.R) {visibility="uncounted" .center data-menu-title="Exercises"}

<center>
<iframe height="390" width= "520" src="https://www.tiktok.com/player/v1/6811647290709757189?&music_info=1&description=1" allow="fullscreen" title="test"></iframe>
</center>

<!-- assumptions tiktoks 'is it me, jesus?' -->

::: footer
<https://www.tiktok.com/@chelseaparlettpelleriti/video/6811647290709757189>  
[Exercises](https://www.ds4ling.jvcasillas.com/slides/05_lm/02_assumptions_diagnostics/index_files/scripts/assumptions_walkthrough_ex.R)
:::






# [References]{.emph} {.final}

::: notes
@wickham2016r
@qass22_ch2
@qass57_ch1
@figueredo_regression
:::

