---
title   : 'Data Science for Linguists'
subtitle : 'The linear model: [Multiple regression]{style="color: #888; font-size: 0.9em;"}'
author   : "Joseph V. Casillas, PhD"
institute: "Rutgers University<mybr>Spring 2025<br>Last update: `r Sys.Date()`"
---

```{r}
#| label: load-helpers
#| echo: false 
#| message: false 
#| warning: false
source(here::here("assets", "scripts", "helpers.R"))
source(here::here("05_lm", "03_mrc", "index_files", "scripts", "mrc.R"))
```


```{r}
#| label: load-part-1
#| child: "part1"
#| eval: true
```

```{r}
#| label: load-part-2
#| child: "part2"
#| eval: true
```




# [References]{.emph} {.final visibility="uncounted"}

::: {#refs}
:::

::: notes
@wickham2016r
@qass22_ch3
@qass57_ch2
@qass57_ch5
@qass50_ch2
@qass50_ch4
@figueredo_regression
@figueredo_inference
@figueredo_selection
:::




# Bonus [- Statistical control in MRC]{color="white"} {.transition visibility="uncounted"}

---

## {background-image="./index_files/img/mrc_math.png" background-size="contain" visibility="uncounted"}

::: notes
We are interested in testing the hypothesis that taller people are better at math. 

A priori we know that this doesn't make sense. 

In a population of children we know that there is a causal relationship 
between age and height and age and math score
(as children get older they get taller; as children get older they get better 
at math)

Thus, in a population of children we would observe a correlation between height 
and math (age is a confound)
:::

---

## {visibility="uncounted"}

[Some data]{.emph .p-font style="font-size: 1.75em; color: #666666;"}

```{r}
#| label: mrc-math
#| comment: ""
str(mrc_ex_data)
sample_n(mrc_ex_data, 10) |> 
  mutate_if(is.numeric, round, digits = 2) |> 
  head(n = 10)
```

---

## {.smaller .center visibility="uncounted"}

::: {.columns}
::: {.column}
```{r}
#| label: mrc-math-height-age-mod
#| echo: true
lm(height ~ age_c, mrc_ex_data) |> 
  summary()
```
:::

::: {.column}
```{r}
#| label: mrc-math-height-age-plot
#| fig-asp: 0.9
mrc_ex_data |> 
  ggplot() + 
  aes(x = age_c + mean(age), y = height_c + mean(height)) + 
  geom_point(alpha = 0.1) + 
  geom_smooth(
    method = lm, formula = "y ~ x", color = "white", size = 4.5
  ) + 
  geom_smooth(
    method = lm, se = F, formula = "y ~ x", size = 2.5
  ) + 
  labs(y = "Height", x = "Age", 
      title = "Height as a function of age") + 
  ds4ling_bw_theme(base_size = 24)
```
:::
:::

---

## {.smaller .center visibility="uncounted"}

::: {.columns}
::: {.column}
```{r}
#| label: mrc-math-score-age-mod
#| echo: true
lm(score ~ age_c, mrc_ex_data) |> 
  summary()
```
:::

::: {.column}
```{r}
#| label: mrc-math-score-age-plot
#| fig-asp: 0.9
mrc_ex_data |> 
  ggplot() + 
  aes(x = age_c + mean(age), y = score) + 
  geom_point(alpha = 0.1) + 
  geom_smooth(
    method = lm, formula = "y ~ x", color = "white", size = 4.5
  ) + 
  geom_smooth(
    method = lm, se = F, formula = "y ~ x", size = 2.5
  ) + 
  labs(y = "Score", x = "Age", 
      title = "Math score as a function of age") + 
  ds4ling_bw_theme(base_size = 24)
```
:::
:::

---

## {.smaller .center visibility="uncounted"}

::: {.columns}
::: {.column}
```{r}
#| label: mrc-math-score-height-mod
#| echo: true
lm(score ~ height_c, mrc_ex_data) |> 
  summary()
```
:::

::: {.column}
```{r}
#| label: mrc-math-score-height-plot
#| fig-asp: 0.9
mrc_ex_data |> 
  ggplot() + 
  aes(x = height_c + mean(height), y = score) + 
  geom_point(alpha = 0.1) + 
  geom_smooth(
    method = lm, formula = "y ~ x", color = "white", size = 4.5
  ) + 
  geom_smooth(
    method = lm, se = F, formula = "y ~ x", size = 2.5
  ) + 
  labs(y = "Score", x = "Height", 
      title = "Math score as a function of height") + 
  ds4ling_bw_theme(base_size = 24)
```
:::
:::

. . .

::: footnote
**What can we do about this**?
:::

---

## {.smaller .center visibility="uncounted"}

::: {.columns}
::: {.column}
```{r}
#| label: mrc-ex-adj-age-mod
#| echo: true
lm(score ~ height_c + age_c, mrc_ex_data) |> 
  summary()
```
:::


::: {.column}
```{r}
#| label: mrc-ex-adj-age-plot
#| fig-asp: 0.9
mrc_ex_data |> 
  filter(age == 12.5) |> 
  ggplot() + 
  aes(x = height_c + mean(height), y = score) + 
  geom_point() + 
  geom_smooth(method = lm, formula = "y ~ x", color = "white", size = 2) + 
  geom_smooth(method = lm, se = F, formula = "y ~ x", size = 0.75) + 
  labs(y = "Score", x = "Height", 
      title = "Score as a function of height (age = 12.5)") + 
  ds4ling_bw_theme(base_size = 20)

```
:::
:::

. . .

::: footnote
**We partial out the spurious relationship between math score <br> and height by conditioning on age**.
:::

---

## {visibility="uncounted"}

```{r}
#| label: mrc-final
#| fig-asp: 0.6
mrc_ex_data |> 
  mutate(age_fct = round(age) |> as.factor()) |> 
  ggplot() + 
  aes(x = height, y = score) + 
  facet_wrap(~ age_fct) + 
  geom_point(alpha = 0.2) + 
  geom_smooth(method = "lm", formula = "y ~ x") + 
  labs(title = "Math score as a function of height for each age") + 
  ds4ling_bw_theme(base_size = 14)
```
